---
title: "Class12"
author: "Yuxuan Jiang A17324184"
format: pdf
toc: true
---

## Section1.

###Proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
## Section4 Population Scale Analysis

>Q13:Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Read the data:
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
Total sample size
```{r}
nrow(expr)

```
Sample size of each genotype:
```{r}
table(expr$geno)
```
Median expression levels of each genotype:
```{r}
tapply(expr$exp,expr$geno,median)
```

>Q14:Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Box-plot:
```{r}
library(ggplot2)
library(ggsignif)
ggplot(expr)+
  aes(geno,exp,fill=geno)+
  geom_boxplot(notch=T)+
  geom_signif(comparisons=list(c("A/A", "A/G"), c("A/A", "G/G"), 
                               c("A/G", "G/G")),
              map_signif_level = TRUE,tip_length = 0.01,step_increase = 0.1)+ 
  theme_bw()
```
According to the analysis, samples with A/A genotype show a significantly higher expression of ORMDL3 compared to samples with G/G or A/G genotype. Therefore, the SNP genotype does effect the expression of ORMDL3.

