---
title: "Class05:Data viz with ggplot"
author: "Yuxuan Jiang PID:A17324184"
format: pdf
toc: True
---

## Background 

There are lots of ways to make figures in R. These include so-called "base R" graphics (e.g.`plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:

```{r}
head(cars)
```
```{r}
plot(cars)
```

First I need to install the package with the command `install.packages()`.

>**N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want!

Every time we want to use one of those "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point()
```

Add a line to better show relationship between two variables:

```{r}
p <- ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point()+
  geom_smooth(method = lm, se=T)+ 
  labs(title ="Stopping distance of old cars", 
       subtitle="Data from the 'cars' object",
       x="Speed (mph)",
       y="Distance (ft)") 
```

```{r}
p
```

```{r}
p+theme_bw()
```


## Gene expression plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


A first version plot

```{r}
ggplot(genes)+
  aes(Condition1, Condition2)+
  geom_point()
```

```{r}
table(genes$State)
```

Version 2 let's color by `State` so we can see the "up" and "down" significant genes compared to all the "unchanging" genes.

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State)+
  geom_point()
```

Version 3 plot, let's modify the default colors to something we like.

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State)+
  geom_point()+
  scale_color_manual(values=c("blue","grey","red"))+
  labs(title="Gene expression changes upon GLP-1 drug",
       x="Control (no drug)", 
       y="Drug")+
  theme_bw()
```

## Going Further

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
head(gapminder,6)
```

Version 1
```{r}
ggplot(gapminder)+
  aes(gdpPercap,lifeExp, col=continent)+
  geom_point(alpha=0.6)
```

Version 2, let's "facet"(i.e.make a separate plot) by continnet rather than the big hot mess above.
```{r}
ggplot(gapminder)+
  aes(gdpPercap,lifeExp, col=continent)+
  geom_point(alpha=0.6)+
  facet_wrap(~continent)
```

## Custom plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up.
`install.package("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)
gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
filter(gapminder, year==2007, country=="Ireland")
filter(gapminder_2007, country=="United States")
```

```{r}
filter(gapminder, year==1977, country=="United States")
filter(gapminder, year==1977, country=="Ireland")
```

>Q.Make a plot comparing 1977 and 2007 for all coutries.

```{r}
gapminder_7707 <- filter(gapminder,year %in% c(1977,2007))
ggplot(gapminder_7707)+
  aes(gdpPercap,lifeExp, col=continent)+
  geom_point(alpha=0.6)+
  facet_wrap(~year)
```

