---
title: "Class18"
author: "Yuxuan Jiang (A17324184)"
format: pdf
toc: true
---

## Background
Pertuissis (whooping cough) is a common lung infection caused by the bacteria B.Pertussis. It can infect anyonebut is most deadly for infants (under 1 year of age).

## CDC tracking data
The CDC tracks the number of Pertussis cases: 

```{r}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
 1926L,1927L,1928L,1929L,1930L,1931L,
 1932L,1933L,1934L,1935L,1936L,
 1937L,1938L,1939L,1940L,1941L,1942L,
 1943L,1944L,1945L,1946L,1947L,
 1948L,1949L,1950L,1951L,1952L,
 1953L,1954L,1955L,1956L,1957L,1958L,
 1959L,1960L,1961L,1962L,1963L,
 1964L,1965L,1966L,1967L,1968L,1969L,
 1970L,1971L,1972L,1973L,1974L,
 1975L,1976L,1977L,1978L,1979L,1980L,
 1981L,1982L,1983L,1984L,1985L,
 1986L,1987L,1988L,1989L,1990L,
 1991L,1992L,1993L,1994L,1995L,1996L,
 1997L,1998L,1999L,2000L,2001L,
 2002L,2003L,2004L,2005L,2006L,2007L,
 2008L,2009L,2010L,2011L,2012L,
 2013L,2014L,2015L,2016L,2017L,2018L,
 2019L,2020L,2021L,2022L,2023L,2024L,2025L),
 cases = c(107473,164191,165418,152003,
 202210,181411,161799,197371,
 166914,172559,215343,179135,265269,
 180518,147237,214652,227319,103188,
 183866,222202,191383,191890,109873,
 133792,109860,156517,74715,69479,
 120718,68687,45030,37129,60886,
 62786,31732,28295,32148,40005,
 14809,11468,17749,17135,13005,6799,
 7717,9718,4810,3285,4249,3036,
 3287,1759,2402,1738,1010,2177,2063,
 1623,1730,1248,1895,2463,2276,
 3589,4195,2823,3450,4157,4570,
 2719,4083,6586,4617,5137,7796,6564,
 7405,7298,7867,7580,9771,11647,
 25827,25616,15632,10454,13278,
 16858,27550,18719,48277,28639,32971,
 20762,17972,18975,15609,18617,
 6124,2116,3044,7063,22538,21996)
       )

```

> Q1.Year vs cases plot

```{r}
library(ggplot2)
ggplot(cdc)+
  aes(year,cases)+
  geom_point()+
  geom_line()
```

> Q2&Q3 Let's add annotation lines for the major milestones of wP vaccination roll-out (1946), the switch to the aP vaccine (1996), COVID-19(2020).

```{r}
library(ggplot2)
ggplot(cdc)+
  aes(year,cases)+
  geom_point()+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept = 1946,col="blue",linetype="dashed")+
  geom_vline(xintercept=1996,col="red",linetype="dashed")+
  geom_vline(xintercept=2020,col="gray",linetype="dashed")
```
The number of the reported cases dropped dramatically after 1946, highly likely to be due to the wP vaccination roll-out. However, the reported cases increased significantly after the roll-out of the newly developed aP vaccine. This may due to the reduced vaccination rate. In addition, the protective effect provided by aP vaccine may fade faster than the wP vaccine. Finally, the development of more sensitive detection techniques may also contribute to the increase of detected cases. 

## Exploring the CMI-PB data

CMI-PB tracks and makes freely available long-term humoral and cellular immune response data for a large number of individuals who received either DTwP or DTaP combination vaccines in infancy followed by Tdap booster vaccinations. This includes complete API access to longitudinal RNA-Seq, AB Titer, Olink, and live cell assay results directly from their website: https://www.cmi-pb.org/


Basically, make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for pertussis infection.

### The CMI-PB API returns JSON data

They make their data available as JSON format API. To read these types of files into R we will use the read_json() function from the jsonlite package. 

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject)
```

> Q4. How many aP and wP individuals are there in this `subject` table?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)? Is this representative of the US population?

```{r}
table(subject$race,subject$biological_sex)
```
Due to the impact of COVID-19, the sample population is not representative of the US population.

### Joining multiple tables

We can now read more tables from the CMI-PB database.

```{r}
library(jsonlite)
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE) 
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE) 
head(specimen)
head(ab_titer)
```
To make sense of all this data we need to merge all these tables together. Only then will you know that a given Ab measurement (from the `ab_tilter` table) was collected on a certain data (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).

We can use `dplyr` and the `*_join()` family of functions to do this.

```{r}
library(dplyr)
meta <- inner_join(subject,specimen)
head(meta)
```

Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data.

```{r}
library(dplyr)
abdata <- inner_join(ab_titer,meta)
head(abdata)
```
> Q. How many different Ab isotype values are in this dataset?

```{r}
table(abdata$isotype)
```

> Q. How many different antigen values are in this dataset?

```{r}
table(abdata$antigen)
```

> Q12.What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
The most recent dataset has almost double the number of rows compared to the 2021 and 2022 datasets, suggesting that more participants were included in the study in 2023 than in 2021 and 2022.


Let's focus on IgG isotype:

```{r}
library(dplyr)
igg <- abdata |>
       filter(isotype=="IgG")
head(igg)
```

Plot for `MFI_normalised` vs. `antigen`:

```{r}
library(ggplot2)
ggplot(igg)+
  aes(MFI_normalised,antigen)+
  geom_boxplot()
```

The antigen "PT", "FIM2/3" and "FHA" appears to have the widest range of values.

> Q. is there a difference for thses responses between aP and wP individual?

```{r}
library(ggplot2)
ggplot(igg)+
  aes(MFI_normalised,antigen,col=infancy_vac)+
  geom_boxplot()
```

```{r}
library(ggplot2)
ggplot(igg)+
  aes(MFI_normalised,antigen,col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

> Q.Is there a difference with time (i.e. before booster shot vs. after booster shot)?

```{r}
library(ggplot2)
ggplot(igg)+
  aes(MFI_normalised,antigen,col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~visit)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```


>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens PT, FIM2/3, and FHA show differences in the level of IgG antibody titers recognizing them over time. Because these antigens are components of the aP vaccine, the observed increase in IgG levels likely reflects the immune activation following vaccination.

```{r}
ab.PT.21 <- abdata |>
             filter(dataset == "2021_dataset",
                    isotype == "IgG",  
                    antigen == "PT")
  ggplot(ab.PT.21) +
    aes(planned_day_relative_to_boost,
        MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line()+
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
ab.PT.21 <- abdata |>
             filter(dataset == "2021_dataset",
                    isotype == "IgG",  
                    antigen == "PT")
  ggplot(ab.PT.21) +
    aes(planned_day_relative_to_boost,
        MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line()+
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")+
    geom_smooth(aes(group = infancy_vac),se=FALSE)
```

```{r}
ab.PT.21 <- abdata |>
             filter(dataset == "2021_dataset",
                    isotype == "IgG",  
                    antigen == "PT")
  ggplot(ab.PT.21) +
    aes(planned_day_relative_to_boost,
        MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")+
    geom_smooth(aes(group = infancy_vac),se=FALSE)
```

